---
title: "Session 4 lab exercise: Poisson log-linear regression"
author: "Levi Waldron"
institute: "CUNY SPH Biostatistics 2"
clean: false
output:
  html_document:
    toc: yes
    df_print: paged
    theme: lumen
    number_sections: yes
  md_document:
    preserve_yaml: false
always_allow_html: true
---

**Learning objectives**

1. Learning objective 1

**Exercises**

1. Example by simulation

```{r, echo=FALSE}
set.seed(1)
```

```{r}
simdat <- data.frame(race=sample(c("white", "non-white"), size=10000, replace=TRUE))
simdat$race <- factor(simdat$race, levels=c("white", "non-white"))
simdat$y <- rpois(10000, lambda=ifelse(simdat$race=="white", exp(3.5), exp(3)))
fit <- glm(y ~ race, data=simdat, family=poisson("log"))
summary(fit)
```

2. Inference on deviance residuals

```{r, echo=FALSE, results='hide'}
(df <- fit$df.residual) #residual degrees of freedom
(res.dev <- sum(fit$deviance))  #total residual deviance
```
* The difference in total deviance between two nested models is $\chi^2$ distributed under $H_0$ that the more complex model is no better at explaining the response.
     + The difference in deviance residuals is (26157 - 10111) = 16046, with a difference of 1 degrees of freedom.

The critical threshold for rejection at p=0.05 is:
```{r}
qchisq(0.95, df=1)
```

So we reject $H_0$

3. Inference on deviance residuals 2

Total residual deviance is $\chi^2$ distributed if the model is correctly specified
     + What is the critical value for rejecting $H_0$ at $p < 0.05$ with a $\chi^2$ distribution of 9998 degrees of freedom?
```{r}
qchisq(0.95, df=9998)
```

Here total residual deviance is 10111, so we do *not* exceed the threshold and do not reject $H_0$ that the model is correctly specified.

4. Interpret deviance Residuals Plots

```{r, echo=FALSE}
par(mfrow=c(2,2))
plot(fit)
```

5. Example: Risky Drug Use Behavior

* Download the "needle_sharing" dataset
* Outcome is # times the drug user shared a syringe in the past month (shared_syr)
* Predictors: sex, ethn, homeless

```{r}
needledat = read.csv("needle_sharing.csv")
summary(needledat$shared_syr)
var(needledat$shared_syr, na.rm=TRUE)
```

5. Table of Risky Drug Use Behavior dataset

6. Plots of Risky Drug Use Behavior

```{r, echo=FALSE}
par(mfrow=c(1,2), cex=2)
hist(needledat$shared_syr, main="")
plot(sort(needledat$shared_syr), pch=".")
```

* There are a _lot_ of zeros - Poisson model is not a good fit

8. Risky Drug Use Behavior: fitting a Poisson model

```{r, echo=FALSE}
needledat$homeless <- factor(needledat$homeless, levels=0:1, 
                             labels=c("no", "yes"))
```

```{r, echo=FALSE}
fit.pois <- glm(shared_syr ~ sex + ethn + homeless, 
           data=needledat, family=poisson(link="log"))
summary(fit.pois)
```

9. Risky Drug Use Behavior: residuals plots

```{r, echo=FALSE, warning=FALSE}
par(mfrow=c(2,2))
plot(fit.pois)
```
