---
title: "Session 4 lab exercise: Poisson log-linear regression"
author: "Levi Waldron"
institute: "CUNY SPH Biostatistics 2"
clean: false
output:
  html_document:
    toc: yes
    df_print: paged
    theme: lumen
    number_sections: yes
  md_document:
    preserve_yaml: false
always_allow_html: true
---

**Learning objectives**

1. Simulate Poisson-distributed data with a relevant covariate
2. Fit a Poisson log-linear GLM
3. Create and interpret diagnostic plots for a log-linear GLM
4. Use analysis of deviance to compare two log-linear GLMs
5. Practice recoding and creating tables and plots


**Exercises**

1. Simulate count data from a Poisson distribution (for example number of hospital visits by persons over 70 in a 3-year period), where:
    a. each person is annotated with "race" as "white" or "non-white"
    b. "white" persons have an average of 3.5 hospital visits during this time period
    c. "non-white" persons have an average of 3 hospital visits

```{r, echo=FALSE}
set.seed(1)
```

```{r}
simdat <- data.frame(race=sample(c("white", "non-white"), size=10000, replace=TRUE))
simdat$race <- factor(simdat$race, levels=c("white", "non-white"))
simdat$y <- rpois(10000, lambda=ifelse(simdat$race=="white", exp(3.5), exp(3)))
```

2. Fit a log-linear Poisson model of count outcomes with "race" as covariate

```{r}
fit <- glm(y ~ race, data = simdat, family = poisson("log"))
summary(fit)
```

3. Use a chi-square test on deviance residuals to test null hypothesis of no relationship between mean hospital visits and race.

```{r, echo=FALSE, results='hide'}
(df <- fit$df.residual) #residual degrees of freedom
(res.dev <- sum(fit$deviance))  #total residual deviance
```

* The difference in total deviance between two nested models is $\chi^2$ distributed under $H_0$ that the more complex model is no better at explaining the response.
     + The difference in deviance residuals is (26157 - 10111) = 16046, with a difference of 1 degrees of freedom.

The critical threshold for rejection at p=0.05 is:
```{r}
qchisq(0.95, df=1)
```

So we reject $H_0$

3. Inference on deviance residuals 2

Total residual deviance is $\chi^2$ distributed if the model is correctly specified
     + What is the critical value for rejecting $H_0$ at $p < 0.05$ with a $\chi^2$ distribution of 9998 degrees of freedom?
```{r}
qchisq(0.95, df=9998)
```

Here total residual deviance is 10111, so we do *not* exceed the threshold and do not reject $H_0$ that the model is correctly specified.

4. Create and discuss standard fit diagnostics plots

```{r, echo=FALSE}
par(mfrow = c(2, 2))
plot(fit)
```

5. Example: Risky Drug Use Behavior

* Download the "needle_sharing" dataset (see Vittinghoff 8.3.1)
* Outcome is # times the drug user shared a syringe in the past month (`shared_syr`)
* Predictors: sex, ethn, homeless

```{r}
library(readr)
needledat = read_csv("needle_sharing.csv")
summary(needledat$shared_syr)
var(needledat$shared_syr, na.rm=TRUE)
```

Some recoding:
```{r, echo=FALSE}
library(dplyr)
needledat_cleaned <-
  mutate(needledat,
    homeless = factor(homeless, levels = 0:1, labels = c("No", "Yes")),
    sex = factor(sex, levels = c("M", "F"), labels = c("Male", "Female")),
    ethnicity = factor(ethn)
  ) %>%
  select(all_of(c("shared_syr", "ethnicity", "sex", "homeless")))
```

5. Create a table of the risky drug use behavior dataset

```{r tableone}
library(tableone)
t1 <- CreateTableOne(data=needledat_cleaned, includeNA = TRUE)
print(t1, missing=TRUE, showAllLevels = TRUE)
```

6. Plots of Risky Drug Use Behavior

a. Create a histogram number of syringe uses

b. Create a scatter plot of number of syringe uses versus rank of number of syringe uses

```{r}
library(ggplot2)
ggplot(needledat, aes(shared_syr)) +
  geom_histogram() +
  labs(title = "Counts of Syringe Sharing Incidents Per Person") +
  xlab("Number of Incidents") +
  ylab("Number of people with that count")
```

```{r}
library(dplyr)
mutate(needledat, rnk = rank(shared_syr, ties.method = "first")) %>%
  ggplot(aes(x = rnk, y = shared_syr)) +
  geom_point() +
  labs(title = "Count vs Rank Count of Syringe Sharing Incidents") +
  xlab("rank of count") +
  ylab("count of syringe sharing")
```

* There are a _lot_ of zeros - Poisson model is not a good fit

8. Fit a Poisson model to the risky drug use behavior dataset anyways

```{r, echo=FALSE}
fit.pois <- glm(shared_syr ~ sex + ethn + homeless,
                data = needledat,
                family = poisson(link = "log"))
summary(fit.pois)
```

9. Create and discuss Poisson model diagnostic plots

```{r, echo=FALSE, warning=FALSE}
par(mfrow = c(2, 2))
plot(fit.pois)
```
